---
title: "Week 8 - Multiple Regression"
author: "Simen Løkken"
format: html
editor: visual
---

## Set environment and load packages

```{r setup}

knitr::opts_chunk$set(
  warning = F,
  message = F
)
```

```{r}

library(dplyr)
library(ggplot2)
library(GGally)
library(readr)
library(janitor)
library(forcats)
library(purrr)
library(patchwork)
library(gglm)
library(broom)

theme_set(theme_minimal())
```

## Part A

First, we load the data and clean the data:

```{r}

alien_data <- read_csv("https://www.math.ntnu.no/emner/ST2304/2024v/Module07/Alien101LabData.csv")

alien_data <- alien_data |> 
  clean_names()
```

### A1.

I can tell from the plots that temperature, rainfall and biomass is normally distributed, can be well explained by a linear function (linearity) and that they positively correlate with each other.

```{r}

ggpairs(alien_data)
```

### A2.

Well, if I were to structure this experiment I would have let one row represent one day and let the temperature be the mean temperature of that day (either whole day or a part of the day, for example from 8-20) and rainfall be total amount of precipitation for that day. Since the data is structured that way here, I assumes it's from the same day.

```{r}

alien_data
```

We can also somewhat investigate this by using our knowledge of weather and climate. If we plot a scatter plot of temperature and rainfall we see that low the temperature is lower when it rains. This makes sense as the sun rays who warms are not passing through the clouds which hold the rain as when it's clear and not raining.

```{r}

alien_data |> 
  ggplot(aes(temperature, rainfall)) +
  geom_point()
```

### A3.

Linear models:

```{r}

mod_1 <- lm(biomass ~ temperature, data = alien_data)
mod_2 <- lm(biomass ~ rainfall, data = alien_data)

```

Model coefficients:

```{r}

mod_list <- list("biomass ~ temperature" = mod_1, "biomass ~ rainfall" = mod_2)

map(mod_list, coef)
```

Model CI's:

```{r}

map(mod_list, confint)
```

How much of the variance is explained by the model?

```{r}

compute_r_squared <- function(model) {
  r_squared <- summary(model)$r.squared
  return(r_squared)
}

map(mod_list, compute_r_squared)
```

### A4.

The results indicate that temperature has a positive effect on growth and that rainfall has a negative effect on growth.

For a unit increase in temperature, biomass increases with 0.407. For a one unit increase in rainfall, biomass decreases with 0.007. However, the R squared are not "good" for either model (or, 0.62 is not bad) and I would not trust these models as they are right now. If I were to give a statement about conditions for ideal growth, it would be a place where it is warm and not rains a lot. The other way around it would have been a cold place with a lot of rain.

For fun, I choose to run some computations on different values of temperature and rainfall to illustrate how the linear function works for each model:

```{r}

predict_y <- function(alpha, beta, x) {
  y <- alpha + beta * x
  return(y)
}
```

Temperature:

```{r}

map_dbl(10:20, ~ predict_y(alpha = mod_1$coefficients[1], beta = mod_1$coefficients[2], .x))
```

Rainfall:

```{r}

rain_vec <- seq(750, 1250, length.out = 10)

map_dbl(rain_vec, ~ predict_y(alpha = mod_2$coefficients[1], beta = mod_2$coefficients[2], .x))
```

### A5.

```{r}

weather_data <- read_csv("https://www.math.ntnu.no/emner/ST2304/2024v/Module07/AustraliaEnvironmentalData.csv")

weather_data <- weather_data |> 
  rename(
    rainfall = Rain,
    temperature = Temp
  )
```

```{r}

temp_preds <- predict(mod_1, newdata = weather_data, interval = "prediction") |> 
  as_tibble()

rain_preds <- predict(mod_2, newdata = weather_data, interval = "prediction") |> 
  as_tibble()
```

### A6.

```{r}

source("https://www.math.ntnu.no/emner/ST2304/2024v/Module07/plotting_function.R")

plotting_function(weather_data, temp_preds[, 1])
```

```{r}

plotting_function(weather_data, rain_preds[, 1])
```

### A7.

I would recommend the company to use its resources on the parts of Australia that corresponds to the dark blue/purple areas, where the model indicates lowest biomass. Central Australia, for example.

## Part B

### A1.

```{r}

mod_3 <- lm(biomass ~ rainfall + temperature, data = alien_data)

coefs <- coef(mod_3)
```

### A2.

The linear equation for this multiple regression would be:

$Y = \alpha + \beta_1x_1 + \beta_2x_2$

We add one more term which is the beta coefficient for the second explanatory variable and the x value we provide it.

### B3.

The effect of temperature has been slightly reduced when including rainfall in the model (beta = 0.407 to beta = 0.363), and the effect of rainfall has been reduced as well (from beta = -0.007 to beta = -0.002).

```{r}

tidy(mod_1)
```

```{r}

confint(mod_1)
```

```{r}

summary(mod_1)$r.squared
```

```{r}

tidy(mod_2)
```

```{r}

confint(mod_2)
```

```{r}

summary(mod_2)$r.squared
```

```{r}

tidy(mod_3)
```

```{r}

confint(mod_3)
```

```{r}

summary(mod_3)$r.squared
```

### B4.

R used maximum likelihood estimation (MLE) to find these estimates. MLE is a an approach where we use those estimates that are most likely given the data we have, by using probability density functions such as dnorm().

### B5.

I would argue that the fit is OK. The residuals are normally distributed, we know from previous plot that and the residuals vs.fitted that linearity assumption is held, and there are no clear outliers.

```{r}

gglm(mod_3)
```

### B6.

```{r}

new_data <- tibble(
  temperature = weather_data$temperature,
  rainfall = weather_data$rainfall
)

temp_rain_preds <- predict(
  mod_3,
  newdata = new_data,
  interval = "prediction"
) |> 
  as_tibble()
```

```{r}

plotting_function(weather_data, temp_rain_preds[, 1])
```

```{r}

plotting_function(weather_data, temp_rain_preds[, 2])
```

```{r}

plotting_function(weather_data ,temp_rain_preds[, 3])
```

### B7.

My recommendation stays the same (use resources on areas with high biomass), but since the model is different, so is my answer. Central Australia is not that important anymore as the new model indicates that northern Australia is more inclined for uncontrolled Alien plant growth.

I have to mention that even though we now have two explanatory variables (rain and temperature), the variance explained by this model is not much better than the model with only temperature. So, I'm not quite sure how much better these predictions are. Luckily, both models indicate that Northern Australia is better than Southern.

## Part C

### C1.

I'm not a biology student so I don't know much about plants, but I'll try to reason around it.

The model indicates that the alien species thrives in warm weather and do not enjoy too much rain. I believe we can say something about the relative influence of rain and temperature in the combined model as the intepretation is as I've understood it to be the effect of temperature when rain is held constant and vice versa. This is what I've based the first sentence on, that the plant enjoy warmth but not rain.

Regarding if the variables we've used in the analysis are enough, I would say no because we've only explained about 60 % of the variance in biomass. There are most likely other factors that counts, such as sunlight, the quality of the soil and so on.

### C2.

We run multiple linear regression as opposed to simple linear regression individually so we can:

1.  Control for confounding variables, i.e., we can control for the influence of rain when we assess the effect of temperature on biomass.
2.  Real-world phenomena are rarely explained by one thing, so if we include several variables we are able to more accurately describe the world.

## Part D

I would like feedback on intepretation on statistical concepts. Overall I think this exercise went great, and I liked the increase in complexity on the coding part. I felt like you did remove some of the hints we got in previous weeks.



