---
title: "Week 5: The Normal Distribution"
author: "Simen Løkken"
format: html
editor: visual
---

## Set environment and load packages

```{r setup}

knitr::opts_chunk$set(
  warning = F,
  message = F
)
```

```{r}

library(dplyr)
library(readr)
library(purrr)
library(ggplot2)
library(janitor)

theme_set(theme_minimal())
```

## Module

### Some important probability remaks

An assumption for doing modelling with several data points is to assume that they are independent of each other, i.e., Pr(x1, x2) = Pr(x1) * Pr(x2).

The likelihood is the probability of each data point as a function of the parameters, we calculate the likelihood for the parameters as a function of the data (Xi's) by multiplying the probabilities.

### Groundhoug day

Read data:

```{r}

groundhog_data <- 
  read_csv("https://www.math.ntnu.no/emner/ST2304/2024v/Module04/GroundhogData.csv")

groundhog_data <- groundhog_data |> 
  clean_names() |> 
  select(-x1)
```

Investigate temperature distribution:

```{r}

groundhog_data |> 
  ggplot(aes(temperature)) +
  geom_histogram(fill = "royalblue", color = "black", alpha = 0.9)
```

We are trying to summarise the data as a distribution so we can ask detailed questions, develop theories about it and use it for predictions with maths and probability from the distribution we believe the data follows.

We often log-transform the normal distribution for easier mathematics.

### Normal distribution exercises

Source functions:

```{r}

source("https://www.math.ntnu.no/emner/ST2304/2024v/Module04/NormalDistFunctions.R")
```

#### 1.

```{r}

mean_temp <- mean(groundhog_data$temperature)
sd_temp <- sd(groundhog_data$temperature)

sim_data <- tibble(
  sims = rnorm(1e3, mean_temp, sd_temp)
)

sim_data |> 
  ggplot(aes(sims)) +
  geom_histogram(fill = "royalblue", color = "black", alpha = 0.9)

```
  
#### 2, 3 & 4.

```{r}

sim_data <- tibble(
  at_data = seq(min(groundhog_data$temperature), max(groundhog_data$temperature), length = 20),
  dens_data = dnorm(at_data, mean_temp, sd_temp),
  cum_data = pnorm(at_data, mean_temp, sd_temp)
)

sim_data |> 
  ggplot(aes(at_data, dens_data)) +
  geom_path()

sim_data |> 
  ggplot(aes(at_data, cum_data)) +
  geom_path()

quants_sim <- tibble(
  quantiles = c(0.025, 0.25, 0.50, 0.75, 0.975),
  quant_data = qnorm(quantiles, mean_temp, sd_temp)
)

quants_sim |> 
  ggplot(aes(quantiles, quant_data)) +
  geom_path()
  
```

#### 5.

The pnorm calculates the cumulative density function which gives the probability of a variable being less than or equal to a value. 

The qnorm is the inverse of the cumulative density function because it takes a vector of probablities as input and gives quantiles as output. It finds the value of our variable at the percentile(s) we provide.

### The log-likelihood

```{r}

calculate_loglikelihood <- function(mu, sigma, data) {
  loglikelihood <- sum(dnorm(data, mu, sigma, log = T))
  return(loglikelihood)
}

sim_data <- tibble(
  means = seq(23, 35, length.out = 500),
  loglikelihoods = map_dbl(
    .x = means,
    .f = ~ calculate_loglikelihood(
      mu = .x, sigma = sd_temp, data = groundhog_data$temperature
    )
  )
)

sim_data |> 
  ggplot(aes(means, loglikelihoods)) +
  geom_path()
```

The code above provides the log likelihood for mu as a function of a single guess for sigma. We see that because the provide a vector of different values for mu, but only one value for sigma.

```{r}

sim_data |> 
  ggplot(aes(means, exp(loglikelihoods - max(loglikelihoods)))) +
  geom_path()
```

### Confidence intervals

First we create a function that takes the mean of N random values from a normal distribution and replicate this n times. 

```{r}

simulate_norm_means <- function(mu, sigma, n, num_repetitions = 10) {
  sims <- replicate(num_repetitions, mean(rnorm(n, mu, sigma)))
  return(sims)
}

sims <- tibble(
  sims = simulate_norm_means(1, 2, 10, 1e4)
)

sims |> 
  ggplot(aes(sims)) +
  geom_histogram(fill = "royalblue", color = "black", alpha = 0.9)

quantile(sims$sims, c(0.025, 0.975))
```

#### 1.

Estimate MLE's for mu and sigma:

```{r}

mu <- mean(groundhog_data$temperature)
var <- var(groundhog_data$temperature)
sigma <- sqrt(var)

temp_length <- length(groundhog_data$temperature)

sims <- tibble(
  sims = simulate_norm_means(mu, sigma, temp_length, num_repetitions = 1e4)
)

sims |> 
  ggplot(aes(sims)) +
  geom_histogram(fill = "royalblue", color = "black", alpha = 0.9)

quantile(sims$sims, c(0.025, 0.975))
```

```{r}

dt(x = 0.5, df = 5)

pt(q = 1.5, df = 5)

qt(p = 0.7, df = 5)

rt(n = 2, df = 5)
```




